---
title: "DS311 - R Lab Assignment"
author: "Andrew Raul"
date: "`r Sys.Date()` 3/22/23"
output: 
  html_document:
    theme: united
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Assignment 1

* In this assignment, we are going to apply some of the build in data set in R for descriptive statistics analysis.
* To earn full grade in this assignment, students need to complete the coding tasks for each question to get the result.
* After finished all the questions, knit the document into HTML format for submission.

### Question 1

Using the **mtcars** data set in R, please answer the following questions.

```{r}
# Loading the data
data(mtcars)

# Head of the data set
head(mtcars)
```

a. Report the number of variables and observations in the data set.

```{r}
# Enter your code here!
ncol(mtcars)
nrow(mtcars)

# Answer:
print("There are total of 11 variables and 32 observations in this data set.")
```



b. Print the summary statistics of the data set and report how many discrete and continuous variables are in the data set.

```{r}
# Enter your code here!
summary(mtcars)

# Answer:
print("There are 7 discrete variables and 4 continuous variables in this data set.")
```



c. Calculate the mean, variance, and standard deviation for the variable **mpg** and assign them into variable names m, v, and s. Report the results in the print statement.

```{r}
# Enter your code here!
m <- mean(mtcars$mpg)
v <- var(mtcars$mpg)
s <- sd(mtcars$mpg)

print(paste("The average of Mile Per Gallon from this data set is ", m , " with variance ", v , " and standard deviation", s , "."))
```



d. Create two tables to summarize 1) average mpg for each cylinder class and 2) the standard deviation of mpg for each gear class.

```{r}
# Enter your code here!
avg_mpg_per_cyl <- aggregate(mpg ~ cyl, data=mtcars, mean)
avg_mpg_per_cyl

sd_mpg_per_gear <- aggregate(mpg ~ gear, data=mtcars, sd)
sd_mpg_per_gear

```



e. Create a crosstab that shows the number of observations belong to each cylinder and gear class combinations. The table should show how many observations given the car has 4 cylinders with 3 gears, 4 cylinders with 4 gears, etc. Report which combination is recorded in this data set and how many observations for this type of car.

```{r}
# Enter your code here!
obs_cyl_gear <- table(mtcars$cyl, mtcars$gear)
obs_cyl_gear
print("The most common car type in this data set is car with 8 cylinders and 3 gears. There are total of 12 cars belong to this specification in the data set.")
```


*** 

### Question 2

Use different visualization tools to summarize the data sets in this question.

a. Using the **PlantGrowth** data set, visualize and compare the weight of the plant in the three separated group. Give labels to the title, x-axis, and y-axis on the graph. Write a paragraph to summarize your findings.

```{r}
# Load the data set
data("PlantGrowth")

# Head of the data set
head(PlantGrowth)

# Enter your code here!
boxplot(weight ~ group, data = PlantGrowth, main = "Plant Growth by Group", xlab = "Group", ylab = "Weight")

ctrlGroup <- subset(PlantGrowth, group=='ctrl')
trt1Group <- subset(PlantGrowth, group=='trt1')
trt2Group <- subset(PlantGrowth, group=='trt2')
summary(ctrlGroup)
summary(trt1Group)
summary(trt2Group)


```

Result:

=> Report a paragraph to summarize your findings from the plot!
The plants in the ctrl group have a median weight of 5.1, a mean of 5, a minimum of 4.1, and a maximum of 6.1. The plants in the trt1 group have a median weight of 4.5, a mean of 4.6, a minimum of 3.5, and a maximum of 6.The plants in the trt2 group have a median weight of 5.4, a mean of 5.5, a minimum of 4.9, and a maximum of 6.3. This shows that the plants in the trt2 group have a consistently higher weight than the rest with the range being overall higher.


b. Using the **mtcars** data set, plot the histogram for the column **mpg** with 10 breaks. Give labels to the title, x-axis, and y-axis on the graph. Report the most observed mpg class from the data set.

```{r}
# Load the data set
data("mtcars")

# Head of the data set
head(mtcars)

# Enter your code here!
hist(mtcars$mpg,breaks = 10, main = "Vehicle MPGs", xlab = "MPG", ylab = "Num Cars")

print("Most of the cars in this data set are in the class of 15 mile per gallon.")
```



c. Using the **USArrests** data set, create a pairs plot to display the correlations between the variables in the data set. Plot the scatter plot with **Murder** and **Assault**. Give labels to the title, x-axis, and y-axis on the graph. Write a paragraph to summarize your results from both plots.

```{r}
# Load the data set
data("USArrests")

# Head of the data set
head(USArrests)

# Enter your code here!
pairs(USArrests$Murder ~ USArrests$Assault, main = "Arrest data")

plot(USArrests$Murder, USArrests$Assault, main = "US arrests for Assault vs Murder", xlab = "Murder", ylab = "Assault")

```

Result:

=> Report a paragraph to summarize your findings from the plot!
From the graphs we can see that there is a somewhat strong positive correlation between the number of arrests for assault and the number of arrests for murder in the United States. As the number of arrests for assault increases, so does the number of arrests for murder.


***

### Question 3

Download the housing data set from www.jaredlander.com and find out what explains the housing prices in New York City.

Note: Check your working directory to make sure that you can download the data into the data folder. 

```{r, echo=FALSE}
# Load and clean the housing data set 
download.file(url='https://www.jaredlander.com/data/housing.csv',
              destfile='data/housing.csv', mode='wb')
housingData <- read.csv('data/housing.csv')
housingData <- subset(housingData, 
                      select = c("Neighborhood", "Market.Value.per.SqFt", "Boro", "Year.Built"))
housingData <- na.omit(housingData)                   
```

a. Create your own descriptive statistics and aggregation tables to summarize the data set and find any meaningful results between different variables in the data set.

```{r}
# Head of the cleaned data set
head(housingData)

# Enter your code here!
#summary stats of the values and years built
summary(housingData$Market.Value.per.SqFt)
summary(housingData$Year.Built)

#average value of homes by neighborhood
avg_val_by_neigborhood <- aggregate(Market.Value.per.SqFt ~ Neighborhood, data = housingData, mean)
avg_val_by_neigborhood

#average value of homes by boro
avg_val_by_boro <- aggregate(Market.Value.per.SqFt ~ Boro, data = housingData, mean)
avg_val_by_boro

#average value of houses by SqFt in Financial neighborhood
houses_in_financial_neighboorhood <- subset(housingData, Neighborhood == "FINANCIAL")
mean_value_per_sqft_financial <-mean(houses_in_financial_neighboorhood$Market.Value.per.SqFt)
mean_value_per_sqft_financial

#average value of houses by SqFt in Manhattan boro
houses_in_manhattan_boro <- subset(housingData, Boro == "Manhattan")
mean_value_per_sqft_manhattan <-mean(houses_in_manhattan_boro$Market.Value.per.SqFt)
mean_value_per_sqft_manhattan

#average home value by date
avg_val_by_date <- aggregate(Market.Value.per.SqFt ~ Year.Built, data = housingData, mean)
avg_val_by_date

```

b. Create multiple plots to demonstrates the correlations between different variables. Remember to label all axes and give title to each graph.

```{r}
# Enter your code here!
#scatterplot of average home values by date
plot(avg_val_by_date$Market.Value.per.SqFt, avg_val_by_date$Year.Built, main = "Average Value of SqFt by Year Built", xlab = "Market Value per SqFt", ylab = "Year Built")

#histogram of number of houses in manhattan by year built
hist(houses_in_manhattan_boro$Year.Built, breaks = 7, main = "Number of houses by year built in Manhattan", xlab = "Year Built", ylab = "Number of Houses")

```

c. Write a summary about your findings from this exercise.

=> Enter your answer here!
From the two summary statistics we can see the range of years that houses were built in all of New York, as well as the average price of all houses in New York. Then we can see the break down of the average house price by boro and neighborhood. Finally we see in the stats the average value by each year of being built. In the graphs we can see the data laid out for the average value by SqFt by year built and how many homes in Manhattan were built in certain years.

